---
title: "Download 990 Efile Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download-data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette describes the raw 990 efile fields that `get_features()` needs and
shows how to assemble them — either with the companion **panel990** package (the
recommended path) or by hand.

## Import with panel990 (recommended)

`get_governance_data()` uses the
[panel990](https://github.com/Nonprofit-Open-Data-Collective/panel990) package to
download the required tables, keep the needed columns, merge them into one row
per filing, and drop 990EZ filers — returning a data frame ready for
`get_features()`. Install panel990 first:

```{r, eval=FALSE}
remotes::install_github("Nonprofit-Open-Data-Collective/panel990")
```

Then import a year (optionally restricting to specific EINs) and score in one
step:

```{r, eval=FALSE}
library(governance)

# Assemble the raw fields for a tax year
dat <- get_governance_data(years = 2022)

features <- get_features(dat)
scores   <- get_scores(features)

# ...or the whole pipeline in one call
scores <- get_governance_scores(years = 2022)
```

By default the data comes from the NCCS efile v2.1 archive. To read tables you
have already downloaded, point the source at a local directory:

```{r, eval=FALSE}
src <- panel990::data_source(root = "path/to/efile/csvs")
dat <- get_governance_data(years = 2022, source = src)
```

The rest of this vignette documents the fields those tables must contain — useful
whether you use panel990 or assemble the input yourself.

## The input contract

`get_features()` expects a data frame with one row per filing and, at minimum,
the identifier `ORG_EIN` plus the raw governance fields from Part IV, Part VI,
Part XII, and Schedule M listed below. Values may be encoded as
`"true"`/`"false"`, `"1"`/`"0"`, `"X"` flags, or blanks (`NA` or `""`);
`get_features()` normalizes all of these.

| Part/Schedule | Variable | 990 line | Question |
|---|---|---|---|
| Part IV | `F9_04_AFS_IND_X` | 12a | Independent audited financial statements? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_X` | 28a | Business transaction with a current/former officer, director, trustee, or key employee (DTK)? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_FAM_X` | 28b | ...with a family member of a DTK? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_ENTITY_X` | 28c | ...with a 35%-controlled entity of a DTK? |
| Part IV | `F9_04_CONTR_NONCSH_MT_25K_X` | 29 | Received > $25,000 in non-cash contributions? |
| Part IV | `F9_04_CONTR_ART_HIST_X` | 30 | Received contributions of art, historical treasures, etc.? |
| Part VI | `F9_06_GVRN_NUM_VOTING_MEMB` | 1a | Number of voting governing body members |
| Part VI | `F9_06_GVRN_NUM_VOTING_MEMB_IND` | 1b | Number of independent voting members |
| Part VI | `F9_06_GVRN_DTK_FAMBIZ_RELATION_X` | 2 | Family/business relationship among officers, directors, trustees, key employees? |
| Part VI | `F9_06_GVRN_DELEGATE_MGMT_DUTY_X` | 3 | Delegation of management duties? |
| Part VI | `F9_06_GVRN_DOC_GVRN_BODY_X` | 8a | Contemporaneous documentation of governing body meetings? |
| Part VI | `F9_06_POLICY_FORM990_GVRN_BODY_X` | 11a | Form 990 provided to governing body? |
| Part VI | `F9_06_POLICY_COI_X` | 12a | Conflict of interest policy? |
| Part VI | `F9_06_POLICY_COI_DISCLOSURE_X` | 12b | Annual disclosure of interests? |
| Part VI | `F9_06_POLICY_COI_MONITOR_X` | 12c | Regular monitoring and enforcement of COI policy? |
| Part VI | `F9_06_POLICY_WHSTLBLWR_X` | 13 | Whistleblower policy? |
| Part VI | `F9_06_POLICY_DOC_RETENTION_X` | 14 | Document retention and destruction policy? |
| Part VI | `F9_06_POLICY_COMP_PROCESS_CEO_X` | 15a | Independent compensation process for CEO/top official? |
| Part VI | `F9_06_DISCLOSURE_AVBL_OTH_X` | 18 | Governing documents available: other |
| Part VI | `F9_06_DISCLOSURE_AVBL_OTH_WEB_X` | 18 | ...on another organization's website |
| Part VI | `F9_06_DISCLOSURE_AVBL_REQUEST_X` | 18 | ...upon request |
| Part VI | `F9_06_DISCLOSURE_AVBL_OWN_WEB_X` | 18 | ...on the organization's own website |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_ACCRU_X` | 1 | Accounting method: accrual |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_CASH_X` | 1 | Accounting method: cash |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_OTH` | 1 | Accounting method: other (free text) |
| Schedule M | `SM_01_REVIEW_PROCESS_UNUSUAL_X` | 31 | Review process for unusual noncash gifts? |

For a full data dictionary see the
[irs990efile data dictionary](https://nonprofit-open-data-collective.github.io/irs990efile/data-dictionary/data-dictionary.html).

## Where the tables live

The NCCS efile v2.1 archive hosts one CSV per Part/Schedule and year at:

```
https://nccs-efile.s3.us-east-1.amazonaws.com/public/efile_v2_1/{TABLE}-{YEAR}.CSV
```

See the [efile v2.1 catalog](https://nccs.urban.org/nccs/catalogs/catalog-efile-v2_1.html)
to browse Parts, Schedules, and available years (2009–2024). The four tables
this package needs are:

- `F9-P04-T00-REQUIRED-SCHEDULES`
- `F9-P06-T00-GOVERNANCE`
- `F9-P12-T00-FINANCIAL-REPORTING`
- `SM-P01-T00-NONCASH-CONTRIBUTIONS`

## Assembling the input manually

The code below pulls the four tables for one year, keeps only the fields needed,
and merges them into a single data frame ready for `get_features()`. It is not
run here (it downloads full-year tables); it mirrors
[`data-raw/make-dat-example.R`](https://github.com/Nonprofit-Open-Data-Collective/governance/blob/main/data-raw/make-dat-example.R),
which builds the bundled `dat_example`.

```{r, eval=FALSE}
library(data.table)
library(dplyr)

ROOT <- "https://nccs-efile.s3.us-east-1.amazonaws.com/public/efile_v2_1/"
YEAR <- 2022

get_table <- function(table.name, select) {
  options(timeout = max(600, getOption("timeout")))
  fread(paste0(ROOT, table.name, "-", YEAR, ".CSV"),
        colClasses = "character", select = select)
}

id <- c("OBJECTID", "ORG_EIN")

p06 <- get_table("F9-P06-T00-GOVERNANCE",
                 c(id, "RETURN_TYPE",
                   "F9_06_GVRN_NUM_VOTING_MEMB", "F9_06_GVRN_NUM_VOTING_MEMB_IND",
                   "F9_06_GVRN_DTK_FAMBIZ_RELATION_X", "F9_06_GVRN_DELEGATE_MGMT_DUTY_X",
                   "F9_06_GVRN_DOC_GVRN_BODY_X", "F9_06_POLICY_FORM990_GVRN_BODY_X",
                   "F9_06_POLICY_COI_X", "F9_06_POLICY_COI_DISCLOSURE_X",
                   "F9_06_POLICY_COI_MONITOR_X", "F9_06_POLICY_WHSTLBLWR_X",
                   "F9_06_POLICY_DOC_RETENTION_X", "F9_06_POLICY_COMP_PROCESS_CEO_X",
                   "F9_06_DISCLOSURE_AVBL_OTH_X", "F9_06_DISCLOSURE_AVBL_OTH_WEB_X",
                   "F9_06_DISCLOSURE_AVBL_REQUEST_X", "F9_06_DISCLOSURE_AVBL_OWN_WEB_X"))

p04 <- get_table("F9-P04-T00-REQUIRED-SCHEDULES",
                 c("OBJECTID",
                   "F9_04_AFS_IND_X", "F9_04_BIZ_TRANSAC_DTK_X",
                   "F9_04_BIZ_TRANSAC_DTK_FAM_X", "F9_04_BIZ_TRANSAC_DTK_ENTITY_X",
                   "F9_04_CONTR_NONCSH_MT_25K_X", "F9_04_CONTR_ART_HIST_X"))

p12 <- get_table("F9-P12-T00-FINANCIAL-REPORTING",
                 c("OBJECTID",
                   "F9_12_FINSTAT_METHOD_ACC_OTH", "F9_12_FINSTAT_METHOD_ACC_ACCRU_X",
                   "F9_12_FINSTAT_METHOD_ACC_CASH_X"))

m01 <- get_table("SM-P01-T00-NONCASH-CONTRIBUTIONS",
                 c("OBJECTID", "SM_01_REVIEW_PROCESS_UNUSUAL_X"))

# OBJECTID is unique per filing. Parts IV/VI/XII are core 990 parts (inner join);
# Schedule M is filed by only some orgs (left join). Drop EZ filers, who lack
# these fields.
dat_govern <- p06 |>
  merge(p04, by = "OBJECTID") |>
  merge(p12, by = "OBJECTID") |>
  merge(m01, by = "OBJECTID", all.x = TRUE) |>
  filter(RETURN_TYPE != "990EZ")
```

The resulting `dat_govern` can be passed straight to `get_features()`; see
`vignette("governance-workflow")`.
