---
title: "Getting 990 Data with panel990"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download-data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## governance and panel990

The **governance** package *scores* nonprofit governance; it does not retrieve
data. Retrieving, normalizing, and assembling IRS 990 efile tables is the job of
the companion
[panel990](https://github.com/Nonprofit-Open-Data-Collective/panel990) package.

| Package | Role |
|---|---|
| **panel990** | Download efile tables, filter to a sample frame, merge to one row per filing, build panels. |
| **governance** | Normalize the governance fields into a feature matrix (`get_features()`) and compute the governance index (`get_scores()`). |

`governance` calls `panel990` for you through `get_governance_data()` and
`get_governance_scores()`, so most users never touch `panel990` directly. This
vignette shows both the one-call path and the underlying `panel990` workflow for
when you want more control. Install `panel990` first:

```{r, eval=FALSE}
remotes::install_github("Nonprofit-Open-Data-Collective/panel990")
```

## Required 990 tables

The governance index is built from four efile tables — Form 990 Parts IV, VI, and
XII and Schedule M. Each is a `panel990` table name you pass to `panelize()`
(Part XII also has the short alias `"P12"`).

| Part / Schedule | Governance information it contains | panel990 table name |
|---|---|---|
| Part IV | Required-schedule triggers: audited financials, interested-person business transactions, large non-cash and art/historical contributions | `F9-P04-T00-REQUIRED-SCHEDULES` |
| Part VI | Governing body composition (voting / independent members) and governance & management policies (conflict-of-interest, whistleblower, document retention, CEO compensation process, document availability) | `F9-P06-T00-GOVERNANCE` |
| Part XII | Financial-reporting method (cash / accrual / other) and audit oversight | `F9-P12-T00-FINANCIAL-REPORTING` |
| Schedule M | Review process for unusual non-cash gifts | `SM-P01-T00-NONCASH-CONTRIBUTIONS` |

> **Full 990 filers only.** The governance index does not apply to 990EZ returns
> — they omit these Parts. Restrict the data to `RETURN_TYPE == "990"` (shown
> below). `get_governance_data()` does this for you.

## Option 1: one call (recommended)

`get_governance_data()` builds the full-990 sample frame, downloads the four
tables, keeps the needed columns, and merges them to one row per filing — ready
for `get_features()`.

```{r, eval=FALSE}
library(governance)

# Assemble the raw fields for a tax year (full 990 filers only)
dat <- get_governance_data(years = 2022)

features <- get_features(dat)
scores   <- get_scores(features)

# ...or the whole pipeline in one call
scores <- get_governance_scores(years = 2022)
```

Restrict to specific organizations with `eins` (formatted `EIN2` values), or
read tables you have already downloaded by pointing the source at a directory:

```{r, eval=FALSE}
scores <- get_governance_scores(years = 2022, eins = c("EIN-43-2031361"))

src <- panel990::data_source(root = "path/to/efile/csvs")
dat <- get_governance_data(years = 2022, source = src)
```

## Option 2: the panel990 workflow

Under the hood, `get_governance_data()` assembles a **sample frame** and calls
`panel990::panelize()`. Doing it yourself gives full control — additional
filters, multiple years, BMF traits, DuckDB backends, and so on.

The `form_type` filter is how you keep only full 990 filers: it is a built-in
sample-frame alias mapping to the `RETURN_TYPE` column, so `form_type = "990"`
keeps only `RETURN_TYPE == "990"`.

```{r, eval=FALSE}
library(panel990)

gov_tables <- c(
  "F9-P04-T00-REQUIRED-SCHEDULES",
  "F9-P06-T00-GOVERNANCE",
  "F9-P12-T00-FINANCIAL-REPORTING",
  "SM-P01-T00-NONCASH-CONTRIBUTIONS"
)

# Sample frame: keep full 990 filers (RETURN_TYPE == "990") via the SFW filter
sfw <- create_sfw("governance panel", form_type = "990")

# Download, filter, and merge the tables into one row per filing
panel <- panelize(sfw = sfw, tables = gov_tables, years = 2022)

dat <- as.data.frame(panel)
scores <- get_scores(get_features(dat))
```

`create_sfw()` accepts other convenience filters too — for example
`state = "CA"` or `years = 2018:2022` — which `panelize()` pushes down to the
download step. See `?panel990::create_sfw` and `?panel990::panelize`.

### A note on normalization

`get_features()` performs the governance-specific normalization itself (blank
checkboxes and counts become `0` regardless of how the source encoded them), so
the tables are imported **raw**.

`panel990::panel_normalize()` only rewrites blank *financial* fields (revenue,
expenses, balance sheet) to zero; it leaves the governance checkbox and count
fields untouched. It is therefore safe to include in a shared pipeline but is
neither required nor sufficient for the governance features — `get_features()`
is what normalizes them. (Do **not** run the general `panel990::normalize()` on
the governance checkboxes before `get_features()`: it converts them to logicals,
which `get_features()` does not expect.)

## The input contract

However you obtain the data, `get_features()` expects a data frame with one row
per filing containing `ORG_EIN` plus the raw fields below. Values may be encoded
as `"true"`/`"false"`, `"1"`/`"0"`, `"X"` flags, or blanks (`NA` or `""`);
`get_features()` normalizes all of these.

| Part/Schedule | Variable | 990 line | Question |
|---|---|---|---|
| Part IV | `F9_04_AFS_IND_X` | 12a | Independent audited financial statements? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_X` | 28a | Business transaction with a current/former officer, director, trustee, or key employee (DTK)? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_FAM_X` | 28b | ...with a family member of a DTK? |
| Part IV | `F9_04_BIZ_TRANSAC_DTK_ENTITY_X` | 28c | ...with a 35%-controlled entity of a DTK? |
| Part IV | `F9_04_CONTR_NONCSH_MT_25K_X` | 29 | Received > $25,000 in non-cash contributions? |
| Part IV | `F9_04_CONTR_ART_HIST_X` | 30 | Received contributions of art, historical treasures, etc.? |
| Part VI | `F9_06_GVRN_NUM_VOTING_MEMB` | 1a | Number of voting governing body members |
| Part VI | `F9_06_GVRN_NUM_VOTING_MEMB_IND` | 1b | Number of independent voting members |
| Part VI | `F9_06_GVRN_DTK_FAMBIZ_RELATION_X` | 2 | Family/business relationship among officers, directors, trustees, key employees? |
| Part VI | `F9_06_GVRN_DELEGATE_MGMT_DUTY_X` | 3 | Delegation of management duties? |
| Part VI | `F9_06_GVRN_DOC_GVRN_BODY_X` | 8a | Contemporaneous documentation of governing body meetings? |
| Part VI | `F9_06_POLICY_FORM990_GVRN_BODY_X` | 11a | Form 990 provided to governing body? |
| Part VI | `F9_06_POLICY_COI_X` | 12a | Conflict of interest policy? |
| Part VI | `F9_06_POLICY_COI_DISCLOSURE_X` | 12b | Annual disclosure of interests? |
| Part VI | `F9_06_POLICY_COI_MONITOR_X` | 12c | Regular monitoring and enforcement of COI policy? |
| Part VI | `F9_06_POLICY_WHSTLBLWR_X` | 13 | Whistleblower policy? |
| Part VI | `F9_06_POLICY_DOC_RETENTION_X` | 14 | Document retention and destruction policy? |
| Part VI | `F9_06_POLICY_COMP_PROCESS_CEO_X` | 15a | Independent compensation process for CEO/top official? |
| Part VI | `F9_06_DISCLOSURE_AVBL_OTH_X` | 18 | Governing documents available: other |
| Part VI | `F9_06_DISCLOSURE_AVBL_OTH_WEB_X` | 18 | ...on another organization's website |
| Part VI | `F9_06_DISCLOSURE_AVBL_REQUEST_X` | 18 | ...upon request |
| Part VI | `F9_06_DISCLOSURE_AVBL_OWN_WEB_X` | 18 | ...on the organization's own website |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_ACCRU_X` | 1 | Accounting method: accrual |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_CASH_X` | 1 | Accounting method: cash |
| Part XII | `F9_12_FINSTAT_METHOD_ACC_OTH` | 1 | Accounting method: other (free text) |
| Schedule M | `SM_01_REVIEW_PROCESS_UNUSUAL_X` | 31 | Review process for unusual noncash gifts? |

For a full data dictionary see the
[irs990efile data dictionary](https://nonprofit-open-data-collective.github.io/irs990efile/data-dictionary/data-dictionary.html).
The tables themselves live in the NCCS efile v2.1 archive
(`https://nccs-efile.s3.us-east-1.amazonaws.com/public/efile_v2_1/{TABLE}-{YEAR}.CSV`,
years 2009–2024); browse them in the
[efile v2.1 catalog](https://nccs.urban.org/nccs/catalogs/catalog-efile-v2_1.html).
