---
title: "Governance Workflow"
author: "Olivia Beck"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# Needed for Functionality
library(governance)
library(dplyr)
# Needed for presentation
library(data.table)
library(ggplot2)
# Needed for additional visualizations
library(stringr)
```

# Governance Workflow

## Step 1: Get Input data 

We assume the user already has the relevant 990 Efile data downloaded. See `vignette("download-data")` for instructions on downloading the need 990 efile data.

We will use a subset of a test set that was already created. See `data-raw/01-get-example-data.R` [here](https://github.com/Nonprofit-Open-Data-Collective/governance/blob/main/data-raw/01-get-example-data.R) for details on how this data set was created. 

```{r}
data("dat_example", package = "governance")

set.seed(57)
keep_rows <- sample(1:nrow(dat_example), 200)
dat_example <- dat_example[keep_rows, ]

```


## Step 2: Get Features Matrix 

Use the `get_features()` function to clean the data and transform it into a features matrix. 

```{r}
features_example <- get_features(dat_example)
head(features_example) %>% data.table()
```


## Step 3: Calculate the Scores 

Use the `get_scores()` function to get the governance scores. 

```{r}
scores_example <- get_scores(features_example)
head(scores_example) %>% data.table()
```

We can visualize the distribution of our governance scores.

```{r}
hist(scores_example$total.score,
     main = "Histogram of Goverance Scores",
     xlab = "Total Score")

```

## Additional Visualizations

We can also join our data with the relevance Business Master File (BMF) data [here](https://nccs.urban.org/nccs/catalogs/catalog-bmf.html) to explore other aspects of the organizations relative to their scores. 

### Loading in the BMF Data

Since our test data set is from 2013, we first load in the 2013 BMF and only keep 
data from the organizations that are in our test set. 

```{r}
# Read in BMF data
link2013 <- "https://nccsdata.s3.us-east-1.amazonaws.com/legacy/bmf/BMF-2013-12-501CX-NONPROFIT-PX.csv"
bmf.2013 <- read.csv(link2013)

#format EIN's correctly 
bmf.2013$EIN <- as.numeric(stringr::str_pad(bmf.2013$EIN, 9, pad = "0"))

# Get needed EIN's 
EIN.keep <- dat_example$ORG_EIN
bmf.keep <- bmf.2013[ bmf.2013$EIN %in% EIN.keep, ]

#join them 
plot_example <- 
  scores_example %>%
  left_join(bmf.keep, by = join_by(ORG_EIN == EIN))


```

We can plot the distribution of scores by various characteristics of the organizations. 

```{r}
plot_example %>% 
  select(total.score, ASSETS) %>% 
  ggplot(aes(y = total.score, x = log(ASSETS))) +
  geom_point() +
  ggtitle("Governance Scores vs. Total Assests") +
  xlab("Total Governance Score") +
  ylab("log(Total Assests)")

```

