---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(governance)
library(dplyr)
```

The **governance** package turns raw IRS Form 990 governance and management
fields into a multi-dimensional governance index for benchmarking nonprofits.
The workflow is two steps: normalize the raw fields into a binary feature matrix
with `get_features()`, then score them against a pre-fit factor model with
`get_scores()`.

## Step 1: Get input data

`get_features()` expects a data frame of raw 990 efile fields — the governance,
management, and disclosure questions from Part IV, Part VI, Part XII, and
Schedule M. See `vignette("download-data")` for the exact fields and where they
come from. Retrieval and management of the efile tables is handled by the
companion **panel990** package (in development); until then you can pull the
tables directly from the
[NCCS efile v2.1 archive](https://nccs.urban.org/nccs/catalogs/catalog-efile-v2_1.html).

The package ships a ready-to-use example: a 5,000-organization sample of raw
2022 fields.

```{r}
data("dat_example", package = "governance")

set.seed(57)
dat_example <- dat_example[sample(seq_len(nrow(dat_example)), 200), ]
```

## Step 2: Build the feature matrix

`get_features()` normalizes the raw fields (yes/no, `"X"` flags, member counts)
into 12 binary governance features, appended to the input.

```{r}
features_example <- get_features(dat_example)

features_example %>%
  select(ORG_EIN, P6_LINE_1, P6_LINE_12_13_14, P12_LINE_1) %>%
  head()
```

## Step 3: Calculate the scores

`get_scores()` applies the pre-fit factor model and appends six factor scores
plus a `total.score`.

```{r}
scores_example <- get_scores(features_example)

scores_example %>%
  select(ORG_EIN, total.score) %>%
  head()
```

```{r, fig.width=6, fig.height=4}
hist(scores_example$total.score,
     main = "Distribution of Governance Scores",
     xlab = "Total Score")
```

See `vignette("governance-workflow")` for a fuller walk-through and
`vignette("making-gov-scores")` for the methodology behind the index.
