---
title: "Governance Workflow"
author: "Olivia Beck"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = F}
library(governance)
library(dplyr)
library(data.table)
```

# Governance Workflow

## 1. Get Input data 

We assume the user already has the relevant 990 Efile data downloaded. Instructions for downloading the need 990 efile data, see \code{vignette("download-data", package = "goveranance")}. 

We will use a subset of a test set that was already created. See `data-raw/01-get-example-data.R` for details on how this data set was created. 

```{r}
data("dat_example", package = "governance")

set.seed(57)
keep_rows <- sample(1:nrow(dat_example), 200)
dat_example <- dat_example[keep_rows, ]

```


## Step 2: Get Features Matrix 

Use the `get_features` function to clean the data and transform it into a features matrix. 

```{r}
features_example <- get_features(dat_example)
head(features_example)
```


## Step 3: Calculate the Scores 

Use the `get_scores` function to get the governance scores. 

```{r}
scores_example <- get_scores(features_example)
head(scores_example)
hist(scores_example$total.score)
```
