---
title: "Governance Workflow"
author: "Olivia Beck"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

```{r, message=FALSE, warning=FALSE}
library(governance)
library(dplyr)
library(ggplot2)
```

# Governance Workflow

This vignette walks through the full path from raw 990 fields to governance
scores and shows a few ways to explore the results.

## Step 1: Get input data

`get_features()` consumes raw 990 efile fields from Part IV, Part VI, Part XII,
and Schedule M. In production these come from the companion
[panel990](https://github.com/Nonprofit-Open-Data-Collective/panel990) package
via `get_governance_data(years = 2022)`; see `vignette("download-data")` for the
field list, the required tables, and the full-990 filter. Here we use the
bundled example, a 5,000-organization sample of raw 2022 fields.

```{r}
data("dat_example", package = "governance")

set.seed(57)
dat_example <- dat_example[sample(seq_len(nrow(dat_example)), 200), ]
```

## Step 2: Build the feature matrix

`get_features()` cleans the raw fields and appends 12 binary governance
features. Blank or missing yes/no responses are treated as `0` regardless of how
the source encoded them (`NA` or `""`), so results do not depend on how the
tables were read.

```{r}
features_example <- get_features(dat_example)

features_example |>
  select(ORG_EIN, starts_with("P6_"), P12_LINE_1) |>
  head()
```

Each feature is an indicator of a good-governance practice. We can look at how
common each practice is in the sample:

```{r, fig.width=7, fig.height=4}
feature_cols <- c("P12_LINE_1", "P4_LINE_12", "P4_LINE_28", "P4_LINE_29_30",
                  "P6_LINE_1", "P6_LINE_11A", "P6_LINE_15A", "P6_LINE_18",
                  "P6_LINE_2", "P6_LINE_3", "P6_LINE_8A", "P6_LINE_12_13_14")

rates <- features_example |>
  summarise(across(all_of(feature_cols), ~ mean(.x == 1, na.rm = TRUE))) |>
  tidyr::pivot_longer(everything(), names_to = "feature", values_to = "rate")

ggplot(rates, aes(x = reorder(feature, rate), y = rate)) +
  geom_col() +
  coord_flip() +
  labs(title = "Share of organizations passing each governance feature",
       x = NULL, y = "Share == 1")
```

## Step 3: Calculate the scores

`get_scores()` applies the pre-fit factor model and appends six factor scores
and a `total.score`.

```{r}
scores_example <- get_scores(features_example)

scores_example |>
  select(ORG_EIN, total.score) |>
  head()
```

```{r, fig.width=6, fig.height=4}
ggplot(scores_example, aes(x = total.score)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Governance Scores",
    x = "Total Score", y = "Count"
  )
```

## Joining external data (optional)

Governance scores are most useful alongside other organizational characteristics.
You can join the scores to the NCCS
[Business Master File (BMF)](https://nccs.urban.org/nccs/catalogs/catalog-bmf.html)
on `ORG_EIN` to relate governance to size, sector, or geography. The snippet
below is illustrative (not run here, as it downloads a large file):

```{r, eval=FALSE}
library(readr)
library(stringr)

# Unified BMF (single file across years)
bmf_url <- paste0(
  "https://nccsdata.s3.us-east-1.amazonaws.com/",
  "bmf/unified/v1.2/UNIFIED_BMF_V1.2.csv"
)
bmf <- read_csv(bmf_url)
bmf$EIN <- str_pad(gsub("\\D", "", bmf$EIN), 9, pad = "0")

plot_data <- scores_example |>
  left_join(bmf, by = c("ORG_EIN" = "EIN"))

ggplot(plot_data, aes(x = log(ASSETS), y = total.score)) +
  geom_point(alpha = 0.4) +
  labs(title = "Governance Scores vs. Total Assets",
       x = "log(Total Assets)", y = "Total Governance Score")
```
