---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=F}
library(governance)
library(dplyr)
library(data.table)
```


## Step 1: Get Input Data 

We assume the user already has the relevant 990 Efile data downloaded. Instructions for downloading the need 990 efile data, see \code{vignette("download-data", package = "goveranance")}. 

We will use a subset of a test set that was already created. See `data-raw/01-get-example-data.R`[here](https://github.com/Nonprofit-Open-Data-Collective/governance/blob/main/data-raw/01-get-example-data.R) for details on how this data set was created. 

```{r}
data("dat_example", package = "governance")

set.seed(57)
keep_rows <- sample(1:nrow(dat_example), 200)
dat_example <- dat_example[keep_rows, ]

```


## Step 2: Get Features Matrix 

Use the `get_features()` function to clean the data and transform it into a features matrix. 

```{r}
features_example <- get_features(dat_example)
head(features_example) %>% data.table()
```


## Step 3: Calculate the Scores 

Use the `get_scores()` function to get the governance scores. 

```{r}
scores_example <- get_scores(features_example)
head(scores_example) %>% data.table()
```

We can explore our governance scores.

```{r}
hist(scores_example$total.score,
     main = "Histogram of Goverance Scores",
     xlab = "Total Score")

```
